\Title

### 2013-08-18 synmut.tex
little changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutations rates and
arbitrary fitness functions can be specified. While the best estimates
for the HIV generation time is roughly two days
\citep{perelson_hiv-1_1996, markowitz_novel_2003}, the generation time
has very little influence on the result and basically sets the unit of
time in which other parameters are measured. For simplicity, we use one
generation per day and have checked that the results obtained are
indistinguishable from simulations with a generation time of two days.

\subsubsection*{Parameters}
In order to simulate HIV, we have to specify several parameters that are
known to various degrees. The mutation rate has been measured using
exogenous LacZ constructs and the average nucleotide substitution rate is 
estimated to be $\mu\approx 2 \times 10^{-5}$ per generation \citep{mansky_lower_1995,
abram_nature_2010}. In our simulations, we vary the mutations rate
from $10^{-5} - 4\times 10^{-5}$ per day. Different virions within an
infected person can recombine their genome by coinfection of the same
cell, followed by template switching during reverse transcription. 
Template switching happens between around ten times per reverse
transcription \citep{levy_dynamics_2004}. The combined recombination rate of template
switching and coinfection has been estimated based on modeling to be around $\rho\approx
10^{-5}$ per nucleotide and day
\citep{neher_recombination_2010,batorsky_estimate_2011}, implying a
coinfection rate of a few percent or less, as has been recently
confirmed experimentally \citep{josefsson_single_2013}. We assume a
template switching rate of $10^{-3}$ per nucleotide and vary the
recombination rate between $5\times 10^{-6}- 5\times 10^{-5}$ by adjusting the
coinfection rate. The population size relevant to evolution in chronic
infection has been hotly debated. The number of cells infected by HIV-1
per day is on the order of $10^{7}$ \citep{perelson_hiv-1_1996}, but the
number of viruses contributing to the next generation might be
considerably smaller due to the burstiness of replication. 
Recent evidence points to a relevant population
size in excess of the inverse mutation rate
\citep{boltz_ultrasensitive_2012}. In fact, the evolutionary dynamics
depends only weakly on the actual value of the population size once the
product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
simulate populations between $N=10^4$ and $5 \times 10^4$. Larger populations
become prohibitively costly to simulate.

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $\alpha$ or a deleterious effect $s_d$ with probability
$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 0.75,
and $s_d$ between $2 \times 10^{-4}$ to $2\times 10^{-2}$ per day.
Mutations at the first and second positions were assigned deleterious 
fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
performed and produced similar results). At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate at which new antibody challenges arise is more difficult to
quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
probably driven by escape. Since the sequence of the C2-V5 region of
gp120  changes at a rate of about 1\% per year
\citep{shankarappa_consistent_1999}, this rate should be on the order of
$k_A=0.005$ per day. We simulate a range from $10^{-3.8}$ to $2\times 10^{-2}$ per day.

@@

Most positions are only transiently variable and variants will either fix or
will be lost -- at least in small samples. Given an SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones; in turn, neutral variants fix more frequently than
deleterious ones. The fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$, while it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

\FIG{aft} shows the time course of the frequencies of all synonymous
SNVs (top) and nonsynonymous SNVs (bottom) observed in the
\shankaregion{} region of \env{} in a chronically HIV-1 infected patient (p10 from
\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
reaching high frequency, very few fix (panel~\ref{fig:aftsyn}); in
contrast, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}). This observation seems at odds with
the assumption of neutrality.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous variants more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were not considered; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral expectation is
shown as a black dashed line. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion{} (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}
and show that synonymous mutations fix less often then expected under neutrality. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we come back below.

@@

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases 
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
stretches form stems. We aligned the sequences from each patient to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
lost (only variants reaching frequencies above 15\%). As shown in
\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, its distribution lies between those of
fixed and lost high-frequency variants.

To test the hypothesis that synonymous variants in \shankaregion{} are lost because they
break stems in the conserved stretches between the V loops, we considered
separately SNVs in V loops and their flanks. The greatest
depression in fixation probability is observed in the conserved stems, while the
V loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
regions between the V loops.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs can reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
and green lines in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
cohort.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

@@

\FIG{simfixpvar} shows simulation results for the fixation probability and the
synonymous diversity for different deleterious effects sizes of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure. To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). We restrict this area up to frequency
$\nu=0.5$ since the high frequency part is rather noisy.
If no fixation happens, this restricted area will be
$-0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.09$
for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
probability of synonymous SNVs decreases from the neutral expectation
($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.125$) as the
fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
deleterious SNVs are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution for many different combinations of
effects of synonymous mutations, $s_d$, and the fraction of synonymous
sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
addition, we vary the mutation rate, recombination rate and population
size (see methods).  Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
the distribution of deleterious effects $s_d$ and neutral fraction
$\alpha$ for which we found $-0.11 < A_\text{syn} < -0.06$ and $0.008 <
P_\text{interm} < 0.015$. This subset of parameters indicates that in
order to be consistent with the data only
a minority of synonymous sites ($\alpha \lesssim 0.2$) can be neutral  and that
the deleterious effects of the remainder is on the order of $s_d \sim
0.002$.

The observed distribution of $\alpha$ and $s_d$ is plausible:
(i) A substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
the majority of SNVs reaching high frequency are neutral and no depression is
observed. (ii) In order to hitchhike, the deleterious effect size has to be 
smaller than the escape rate, otherwise the double mutant (with both the escape
mutation and the deleterious synonymous one) has little or no fitness advantage
over the wildtype virus. (iii) The inverse of $s_d$ is of the same order of magnitude as the
fixation/extinction times estimated above (see \FIG{fixp1}). (iv)
The cost $s_d$ tends to be larger than $0.001$, since mutations with smaller
cost behave neutrally and often go to fixation
as expected from the typical coalescent times observed in HIV-1.

The above simulations show that hitchhiking on favored nonsynonymous
variants can reproduce the observed pattern of deleterious synonymous SNVs that
rarely fix. While our model is already quite complicated with many
poorly known parameters that make interpretation of simulation results
challenging, we know that the actual evolutionary dynamics within an
infected person is much more complicated still. In our model, 
 escape mutations are unconditionally beneficial and almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the observed fixation probability of
nonsynonymous SNVs at high frequency often disappear again
(\FIG{fixp2}). One possible reason for this discrepancy is the fact that
the fitness landscape in our model is static: Once a site is designated
as targeted by the immune system, mutations at this site are beneficial
forever. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

@@

Selection against the majority of synonymous substitutions is probably
common across the genome, but we only observe deleterious synonymous SNVs
at high frequency in \shankaregion{} of \env{}, where they hitchhike to
high frequencies on nAb escape mutations. Surprisingly, nonsynonymous
mutations display a fixation pattern as if they were neutral;
see~\FIG{fixp2}. However, nonsynonymous diversity exceeds
synonymous diversity despite the overall much greater constraints on the
amino acid sequence, suggesting that the majority of high frequency SNVs
are escape mutations despite the fact that are often lost again. We suggest that this
paradoxical behavior could be due to (i) escape
mutations that revert after they themselves are recognized by nAbs, or (ii)
the competition between different escape mutations within one epitope. 
Both mechanisms reduce the overall fixation probability and can
give rise to the observed pattern of fixation in computer
simulations. More data is required to investigate this behavior further.

The observed hitchhiking highlights the importance of linkage due to
infrequent recombination for the evolution of HIV-1
\citep{neher_recombination_2010, batorsky_estimate_2011,
josefsson_majority_2011}. The recombination rate has been estimated to be on the
order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
\epsilon^{-1} \log \nu_0$ generations for an escape SNV with escape rate
$\epsilon$ to rise from an initially low frequency $\nu_0$ to frequency
one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
\rho \log \nu_0$ remains linked to the adaptive mutation. With
$\epsilon=0.01$, and $\nu_0\approx 10^{3}$, 
we have $l\approx 100$ bases. Hence we expect strong linkage between the
variable loops and the flanking sequences, but none far beyond the variable
regions, consistent with the lack of signal outside of \shankaregion. In case of
much stronger selection -- such as observed during early CTL escape or drug
resistance evolution -- the linked region is of course much larger
\citep{nijhuis_stochastic_1998}.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.6\linewidth]
{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=0.6\linewidth]
{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) single nucleotide variants (SNVs) in \env, 
\shankaregion, from patient p10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous SNVs fix, few synonymous
SNVs do so even though they are frequently observed at high
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
SNV that reached 50\% frequency is one half.}
\label{fig:aft}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using FFPopSim
\citep{zanini_ffpopsim:_2012}. Briefly, FFPopSim enables individual-based
simulations where each site in the genome is represented by one bit that can be
in one of two states. Outcrossing rates, crossover rates, mutation rates and
arbitrary fitness functions can be specified. While the best estimates
for the HIV generation time is roughly two days
\citep{perelson_hiv-1_1996, markowitz_novel_2003}, the generation time
has very little influence on the result and basically sets the unit of
time in which other parameters are measured. For simplicity, we use one
generation per day and have checked that the results obtained are
indistinguishable from simulations with a generation time of two days.

\subsubsection*{Parameters}
In order to simulate HIV, we have to specify several parameters that are
known to various degrees. The mutation rate has been measured using
exogenous LacZ constructs and the average nucleotide substitution rate is 
estimated to be $\mu\approx 2 \cdot 10^{-5}$ per generation \citep{mansky_lower_1995,
abram_nature_2010}. In our simulations, we vary the mutations rate
from $10^{-5} - 4\cdot 10^{-5}$ per day. Different virions within an
infected person can recombine their genome by coinfection of the same
cell, followed by template switching during reverse transcription. 
Template switching happens between around ten times per reverse
transcription \citep{levy_dynamics_2004}. The combined recombination rate of template
switching and coinfection has been estimated based on modeling to be around $\rho\approx
10^{-5}$ per nucleotide and day
\citep{neher_recombination_2010,batorsky_estimate_2011}, implying a
coinfection rate of a few percent or less, as has been recently
confirmed experimentally \citep{josefsson_single_2013}. We assume a
template switching rate of $10^{-3}$ per nucleotide and vary the
recombination rate between $5\cdot 10^{-6}- 5\cdot 10^{-5}$ by adjusting the
coinfection rate. The population size relevant to evolution in chronic
infection has been hotly debated. The number of cells infected by HIV-1
per day is on the order of $10^{7}$ \citep{perelson_hiv-1_1996}, but the
number of viruses contributing to the next generation might be
considerably smaller due to the burstiness of replication. 
Recent evidence points to a relevant population
size in excess of the inverse mutation rate
\citep{boltz_ultrasensitive_2012}. In fact, the evolutionary dynamics
depends only weakly on the actual value of the population size once the
product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
simulate populations between $N=10^4$ and $5 \cdot 10^4$. Larger populations
become prohibitively costly to simulate.

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $\alpha$ or a deleterious effect $s_d$ with probability
$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 0.75,
and $s_d$ between $2 \cdot 10^{-4}$ to $2\cdot 10^{-2}$ per day.
Mutations at the first and second positions were assigned deleterious 
fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
performed and produced similar results). At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate $k_A$ at which new antibody challenges arise is more difficult to
quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
probably driven by escape. We simulate a range from $10^{-3.8}$ to $2\cdot 10^{-2}$ per day.

@@

Most positions are only transiently variable and variants will either fix or
will be lost -- at least in small samples. Given an SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones; in turn, neutral variants fix more frequently than
deleterious ones. The fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$, while it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aft}A). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

\FIG{aft} shows the time course of the frequencies of all synonymous
SNVs (top) and nonsynonymous SNVs (bottom) observed in the
\shankaregion{} region of \env{} in a chronically HIV-1 infected patient (p10 from
\citet{shankarappa_consistent_1999}). Despite many synonymous SNVs
reaching high frequency, very few fix (Fig.~\ref{fig:aft}A); in
contrast, many nonsynonymous mutations fix
(Fig.~\ref{fig:aft}B). This observation seems at odds with
the assumption of neutrality.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous variants more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were not considered; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (Fig.~\ref{fig:fixp}A). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in Fig.~\ref{fig:fixp}B. The neutral expectation is
shown as a black dashed line. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion{} (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}
and show that synonymous mutations fix less often then expected under neutrality. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we come back below.

@@

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases 
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
stretches form stems. We aligned the sequences from each patient to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
lost (only variants reaching frequencies above 15\%). As shown in
\FIG{SHAPE}A, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, its distribution lies between those of
fixed and lost high-frequency variants.

To test the hypothesis that synonymous variants in \shankaregion{} are lost because they
break stems in the conserved stretches between the V loops, we considered
separately SNVs in V loops and their flanks. The greatest
depression in fixation probability is observed in the conserved stems, while the
V loops show little deviation from the neutral signature, see
\FIG{SHAPE}B. This is suggestive of important RNA helices in conserved
regions between the V loops.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs can reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
and green lines in \FIG{fixp}B).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
cohort.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp}A, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

@@

\FIG{simheat}A shows simulation results for the fixation probability and the
synonymous diversity for different deleterious effects sizes of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure. To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simheat}A, lower inset). We restrict this area up to frequency
$\nu=0.5$ since the high frequency part is rather noisy.
If no fixation happens, this restricted area will be
$-0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.09$
for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
changes. In the three simulations shown in \FIG{simheat}A, the fixation
probability of synonymous SNVs decreases from the neutral expectation
($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.125$) as the
fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
deleterious SNVs are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution for many different combinations of
effects of synonymous mutations, $s_d$, and the fraction of synonymous
sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
addition, we vary the mutation rate, recombination rate and population
size (see methods).  Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Figure \ref{fig:simheat}B shows
the distribution of deleterious effects $s_d$ and neutral fraction
$\alpha$ for which we found $-0.11 < A_\text{syn} < -0.06$ and $0.008 <
P_\text{interm} < 0.015$. This subset of parameters indicates that in
order to be consistent with the data only
a minority of synonymous sites ($\alpha \lesssim 0.2$) can be neutral  and that
the deleterious effects of the remainder is on the order of $s_d \sim
0.002$.

The observed distribution of $\alpha$ and $s_d$ is plausible:
(i) A substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
the majority of SNVs reaching high frequency are neutral and no depression is
observed. (ii) In order to hitchhike, the deleterious effect size has to be 
smaller than the escape rate, otherwise the double mutant (with both the escape
mutation and the deleterious synonymous one) has little or no fitness advantage
over the wildtype virus. (iii) The inverse of $s_d$ is of the same order of magnitude as the
fixation/extinction times estimated above (see \FIG{fixp}A). (iv)
The cost $s_d$ tends to be larger than $0.001$, since mutations with smaller
cost behave neutrally and often go to fixation
as expected from the typical coalescent times observed in HIV-1.

The above simulations show that hitchhiking on favored nonsynonymous
variants can reproduce the observed pattern of deleterious synonymous SNVs that
rarely fix. While our model is already quite complicated with many
poorly known parameters that make interpretation of simulation results
challenging, we know that the actual evolutionary dynamics within an
infected person is much more complicated still. In our model, 
 escape mutations are unconditionally beneficial and almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the observed fixation probability of
nonsynonymous SNVs at high frequency often disappear again
(\FIG{fixp}B). One possible reason for this discrepancy is the fact that
the fitness landscape in our model is static: Once a site is designated
as targeted by the immune system, mutations at this site are beneficial
forever. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

@@

Selection against the majority of synonymous substitutions is probably
common across the genome, but we only observe deleterious synonymous SNVs
at high frequency in \shankaregion{} of \env{}, where they hitchhike to
high frequencies on nAb escape mutations. Surprisingly, nonsynonymous
mutations display a fixation pattern as if they were neutral;
see~\FIG{fixp}B. However, nonsynonymous diversity exceeds
synonymous diversity despite the overall much greater constraints on the
amino acid sequence, suggesting that the majority of high frequency SNVs
are escape mutations despite the fact that are often lost again. We suggest that this
paradoxical behavior could be due to (i) escape
mutations that revert after they themselves are recognized by nAbs, or (ii)
the competition between different escape mutations within one epitope. 
Both mechanisms reduce the overall fixation probability and can
give rise to the observed pattern of fixation in computer
simulations. More data is required to investigate this behavior further.

The observed hitchhiking highlights the importance of linkage due to
infrequent recombination for the evolution of HIV-1
\citep{neher_recombination_2010, batorsky_estimate_2011,
josefsson_majority_2011}. The recombination rate has been estimated to be on the
order of $\rho = 10^{-5}$ per base and day. It takes roughly $t_{sw} =
\epsilon^{-1} |\log \nu_0|$ generations for an escape SNV with escape rate
$\epsilon$ to rise from an initially low frequency $\nu_0$ to frequency
one. This implies that a region of length $l = (\rho t_{sw})^{-1} = \epsilon /
|\rho \log \nu_0|$ remains linked to the adaptive mutation. With
$\epsilon=0.01$, and $\nu_0\approx 10^{-3}$, 
we have $l\approx 100$ bases. Hence we expect strong linkage between the
variable loops and the flanking sequences, but none far beyond the variable
regions, consistent with the lack of signal outside of \shankaregion. In case of
much stronger selection -- such as observed during early CTL escape or drug
resistance evolution -- the linked region is of course much larger
\citep{nijhuis_stochastic_1998}.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.6\linewidth]{fig1}
%\subfloat{\includegraphics[width=0.6\linewidth]
%{Shankarappa_allele_freqs_trajectories_syn_p10.pdf}
%\label{fig:aftsyn}}\\
% \subfloat{\includegraphics[width=0.6\linewidth]
%{Shankarappa_allele_freqs_trajectories_nonsyn_p10.pdf}
%\label{fig:aftnonsyn}}
\caption{Time series of frequencies
of synonymous (A) and nonsynonymous (B) single nucleotide variants (SNVs) in \env, 
\shankaregion, from patient p10~\cite{shankarappa_consistent_1999}.
While many nonsynonymous SNVs fix, few synonymous
SNVs do so even though they are frequently observed at high
frequencies. Colors indicate the position of the site along the \shankaregion{} region
(blue to red). Inset: the fixation probability $\pfix$ of a neutral
SNV that reached 50\% frequency is one half.}
\label{fig:aft}
\end{center}
\end{figure}
